Staged Abstract Interpreters

### 2019-09-11 paper/30_stageconc.tex
polish

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate
on staged values.  For brevity, we call them \textit{staged monads}. In the
code snippet, we also use a @Rep@ prefix on the name of constructors and types
to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@ are not
staged, and that the monadic computation, such as functions of @R => M[A]@, are
also not staged. Instead, the internal data that these monads operate on are
staged, i.e., values of type @R@ and @A@ in the reader monad, and values of type @S@ and
@A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added.
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

@@

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not as straightforward for the nondeterminism monad
(@SetT@) in the abstract interpreter, because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. Therefore, any traversals over the set has to be
staged.  In this section, we explicitly distinguish the two versions of monadic
interfaces. Later, we will instantiate the staged set monad by manually fusing
the fragment of the monad stack inside @SetT@.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$-terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$-term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is passed as @ev@ to the method @close@.
The following code implements the idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

In @close@, we first create a function @f@ that takes a next-stage value and
store, and returns staged values and stores. The function @f@ is a
current-stage and we then use @emit_complied_clo@ to denote it to a next-stage
Scala function.
The method @emit_compiled_clo@ produces current-stage representations of
next-stage function values, i.e., of type @Rep[Value]@.
Inside function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also with the prepared environment and
store.  By doing so, we have \textit{collapsed} the monadic values of @AnsM@ to
grounded values within staging; therefore in the future stage, there is no
monads.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@. But what we can do with these values of type @Rep@ is limited. 
In fact, @fun@ does not have an intensional application operation we can use
directly. We only know that it is a current-stage representation of functions produced by @close@,
whose next-stage form is a Scala function.
With this fact, we can generate a next-stage function application for it, also
in our target language Scala. The method @emit_ap_clo@ produces a current-stage
representation of a next-stage function application, with necessary arguments
such as the latest store @σ@. Meanwhile, we still want the method @ap_clo@
implemented in monadic style, so that it can be smoothly connected with other
parts of the interpreter.  To achieve this, we lift the result values and
stores from the future stage into our current-stage monad stack, with a binding
@vs@.  Finally, we use @put_store@ to update the store with @vs._2@ and return
the value result @vs._1@ into the current-stage monadic value.

Now, compared with the unstaged interpreter, we observe a key difference of where
we use the evaluation function @ev@.  In the unstaged interpreter, the
invocation of @ev@ happens in @ap_clo@, i.e., at the application time; while in
the staged interpreter, we eagerly apply @ev@ when denoting the $\lambda$-terms to
our next-stage value domains, i.e., at the value-representation time.

@@

In the previous section, we showed the types of @emit_compiled_clo@ and
@emit_ap_clo@ without concrete implementations. In this section, we briefly
discuss the IR nodes generated by them and sketch the code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage value definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

>>>>

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that now the monads operate
on staged values.  For brevity, we call them \textit{staged monads}. In the
code snippet, we also use a @Rep@ prefix on the name of constructors and types
to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@ are not
staged, and that the monadic computation, such as functions of @R => M[A]@, are
also not staged. Instead, the internal data that these monads operate on are
staged, i.e., values of type @R@ and @A@ in the reader monad, and values of type @S@ and
@A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added.
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

@@

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not as straightforward for the nondeterminism monad
(@SetT@) in the abstract interpreter, because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements are in the set. Therefore, any traversals over the set has to be
staged.  In this section, we explicitly distinguish the two versions of monadic
interfaces. Later, we will instantiate the staged set monad by manually fusing
the fragment of the monad stack inside @SetT@.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$-terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$-term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is passed as @ev@ to the method @close@.
The following code implements this idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

In @close@, we first create a function @f@, which input and output are both
next-stage values and stores.  The function @f@ is a current-stage and we then
use @emit_complied_clo@ to denote it to a next-stage Scala function.
The method @emit_compiled_clo@ produces current-stage representations of
next-stage function values, i.e., of type @Rep[Value]@.
Inside function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also with the prepared environment and
store.  By doing so, we have \textit{collapsed} the monadic values of @AnsM@ to
grounded values within staging; therefore in the future stage, there will be no
monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@. But what we can do with these values of type @Rep@ is limited. 
In fact, @fun@ does not have an intensional application operation we can use
directly. We only know that it is a current-stage representation of functions produced by @close@,
whose next-stage form is a Scala function.
With this knowledge, we can generate a next-stage function application for it. 
The method @emit_ap_clo@ produces a current-stage
representation of a next-stage function application, with required arguments
such as the latest store @σ@. Meanwhile, we still want the method @ap_clo@
implemented in monadic style, so that it can be smoothly connected with other
parts of the interpreter.  To achieve this, we lift the result values and
stores from the future stage into our current-stage monad stack, with a binding
@vs@.  Finally, we use @put_store@ to update the store with @vs._2@ and return
the value result @vs._1@ into the current-stage monadic value.

Now, compared with the unstaged interpreter, we observe a key difference where
we use the evaluation function @ev@.  In the unstaged interpreter, the
invocation of @ev@ happens in @ap_clo@, i.e., at the time of application; while in
the staged interpreter, we eagerly apply @ev@ when denoting the $\lambda$-terms to
our next-stage value domains, i.e., at the time of value-representation.

@@

In the previous section, we showed the types of @emit_compiled_clo@ and
@emit_ap_clo@ without their concrete implementations. In this section, we briefly
discuss the IR nodes generated by them and sketch the code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage value definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

### 2019-09-11 paper/40_absinterp.tex
misc
polish

After seeing the unstaged and staged concrete interpreter, now we turn our
focus to abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains, such as power sets.  Then, we show the abstract
components, namely, abstract values, abstract environments, and abstract stores.
The abstract interpreter we construct in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For simplicity, it is
context/path/flow-insensitive by our choice. In Section \ref{cfa}, we will
instantiate it to a commonly-used context-sensitive analysis.
\citet{Darais:2015:GTM:2814270.2814308} showed how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our
unstaged or staged abstract interpreter.

@@

The abstract domains operated by abstract interpreters usually can be
formulated by complete lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to the @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to the
trait for annotating the binding-times, where @R[_]@ wraps the data type
@S@ and @Boolean@ in these operations.

@@

In this section, we simply instantiate @SPLattice@ with the flat binding-time
type @R[T] = T@ (i.e., @NoRep@); the result is the trait @Lattice[S]@.
An example of the lattices we used in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}). For two power sets,
they are ordered by the subset relation. We use set union to compute their
join, and set intersection to compute their meet.  The bottom element of a
power set is empty set, and we do not have a representation of top element for
power set.  Other lattices used in the paper, such as products and maps, can be
implemented similarly or by lifting the existing lattices element-wise or
point-wise.  Non-relational numerical abstract domains such intervals can also
be implemented in a stage polymorphic way.

@@

We first widen the concrete value domain to @AbsValue@. There are two variants of
@AbsValue@: 1) numbers are lifted to a singleton object @IntTop@, which represents
the set of all numbers; 2) closures remain the same. Then, the type @Value@ is
redefined as a set of @AbsValue@ to account for approximation.  The address
space is constrained to be finite to ensure the computability of the analysis --
for a simple monovariant analysis, we directly use variable names for
addresses. After defining the @Value@ and @Addr@ type, the types of
environments and stores are automatically lifted to their abstract versions,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@, respectively.  
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ... }
\end{lstlisting}

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore both of the branches.
Also, at some later point, we need to compute the join of two path.  Another
nondeterminism comes from function applications, for instance, @f(a)@, the
possible $\lambda$-terms of @f@ are computed approximately.  As the
result, we need to store all the possible callees, as well as we will
retrieve a set of closures when dereferencing an address.

@@

\paragraph{Monads for Abstract Interpretation}
Compared with the concrete interpreter that uses reader and state effects, the
abstract interpreter further introduces the nondeterminism effect and another
reader and state effect for the two caches. The nondeterminism effect is
represented by the @Set[M[_], A]@ monad transformer, where @M@ is the inner
monad type being transformed, and @A@ is the type of elements in the set. 
We use a @ReaderT@ for one cache that is not changed during one fixed-point
iteration, and use a @StateT@ for another cache that will be constantly updated
during the analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutations of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ... }
\end{lstlisting}

@@

This monad stack scheme (@AnsM[T]@) leads to a different type of grounded
result. Recall that in the concrete setting, we only have a reader monad (for
environments) and a state monad (for stores). The reader effect is not
persistent to the final result. Hence, together with the value produced by the
interpreter, the final result type is just a pair of @Value@ and @Store@.
However, under the new monad stack for abstract interpretation, we have a
@SetT@ monad transformer inside the environment monad and store monad. Therefore, the type
@Set@ becomes the container type of the pairs of values and stores, i.e.,
@Set[(Value, Store)]@.  Also, note that the reader and state monad for the
caches both inhibit internally of the nondeterminism monad; as the result, the
final result type pairs the set of value-stores with the cache type from
that state monad, as shown below.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed according
to the new monad stack scheme and value domains. One of the notable changes is
that the store update operator merges the new values with existing values.  As
we mentioned before, sometimes we have to explore the both branches for
conditionals: in method @br0@, we combine the results using the @mplus@
operation from @MonadPlus@, which requires the value domains are
join-semilattices (in our case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument of @ap_clo@ is a set of closures
@funs@, so we can not directly destruct it to a syntactic @Lam@ by pattern
matching. Instead, we use an auxiliary function @lift_nd@ that takes a set and
lifts the elements in the set into the monad stack. Then we can
straightforwardly implement the @ap_clo@ in monadic style, where the closures
come from the monads and thus the nondeterminism can be naturally handled. The
light gray line shows what is added from the concrete @ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

\paragraph{Caching and Fixpoint Iteration}
As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called a \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use an @in@ cache and an @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and @out@ cache is used for updating in the current iteration. In
the next iteration, the last @out@ cache will be served as the @in@ cache, and
an empty cache is plugged-in to the @out@ slot. Once the @out@ cache does not
contain any new information compared with the @in@ cache, the fixed-point is
reached.
In our monad stack, the @in@ and @out@ caches are modeled by the reader monad and
state monad, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

>>>>

After seeing the unstaged and staged concrete interpreter, we now turn our
focus to abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains, such as power sets.  Then, we show the abstract
components, namely, abstract values, abstract environments, and abstract stores.
The abstract interpreter we construct in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For simplicity, it is
context/path/flow-insensitive by our choice. In Section \ref{cfa}, we will
instantiate it to a commonly-used context-sensitive analysis.
\citet{Darais:2015:GTM:2814270.2814308} showed how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our
unstaged or staged abstract interpreter.

@@

Many abstract domains in abstract interpreters can be formulated by complete
lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage-polymorphic lattices. Similar to the @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to the
trait for annotating the binding-times, where @R[_]@ wraps the data type
@S@ and @Boolean@ in these operations.

@@

In this section, we instantiate @SPLattice@ with the flat binding-time
type @R[T] = T@ (i.e., @NoRep@); the result is the trait @Lattice[S]@.
An example of the lattices we use in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}). Two power sets are
ordered by the subset relation. We use set union to compute their
join, and set intersection to compute their meet.  The bottom element of a
power set is the empty set, and we do not have a representation of the top element for
power set.  Other lattices used in the paper, such as products and maps, can be
implemented similarly or by lifting the existing lattices element-wise or
point-wise.  Non-relational numerical abstract domains such intervals can also
be implemented in a stage-polymorphic way.

@@

We first widen the concrete value domain to @AbsValue@. There are two variants of
@AbsValue@: 1) numbers are lifted to a singleton object @IntTop@, which represents
the set of all numbers; 2) closures remain the same. Then, the type @Value@ is
redefined as a set of @AbsValue@ to account for approximation.  The address
space is constrained to be finite to ensure the computability of the analysis --
for a simple monovariant analysis, we directly use variable names as
addresses. After defining the @Value@ and @Addr@ type, the types of
environments and stores are automatically lifted to their abstract versions,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@, respectively.  
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ... }
\end{lstlisting}

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore both of the branches.
Also, at some later point, we need to compute the join of two paths.  Another
source of nondeterminism comes from function applications, for instance,
consider an application @f(a)@, the possible $\lambda$-terms of @f@ are
computed approximately.  As the result, we need to store all the possible
callees, as well as we will retrieve a set of closures when dereferencing an
address.

@@

\paragraph{Monads for Abstract Interpretation}
Compared with the concrete interpreter that uses reader and state effects, the
abstract interpreter further introduces the nondeterminism effect and another
reader and state effect for the two caches. The nondeterminism effect is
represented by the @Set[M[_], A]@ monad transformer, where @M@ is the inner
monad type being transformed, and @A@ is the type of elements in the set. 
We use a @ReaderT@ for one cache that is not changed during one fixed-point
iteration, and use a @StateT@ for another cache that will be constantly updated
during the analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutations of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what has been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ... }
\end{lstlisting}

@@

This monad stack scheme (@AnsM[T]@) leads to a different type of grounded
values. Recall that in the concrete setting, we only have a reader monad (for
environments) and a state monad (for stores). The reader effect is not
persistent to the final result. Hence, together with the value produced by the
interpreter, the final result type is just a pair of @Value@ and @Store@.
However, under the new monad stack for abstract interpretation, we have a
@SetT@ monad transformer inside the environment monad and store monad. Therefore, the type
@Set@ becomes the container type of the pairs of values and stores, i.e.,
@Set[(Value, Store)]@.  Also, note that the reader and state monad for the
caches both inhibit internally of the nondeterminism monad; as the result, the
final result type pairs the set of value-stores with the cache type from
that state monad, as shown below.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed according
to the new monad stack scheme and value domains. One of the notable changes is
that the store update operator merges the new values with existing values.  As
we mentioned before, sometimes we have to explore both of the branches for
conditionals: in method @br0@, we combine the results using the @mplus@
operation from @MonadPlus@, which requires the value domains are
join-semilattices (in our case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument of @ap_clo@ is a set of closures
@funs@, so we cannot directly destruct it with a syntactic @Lam@ by pattern
matching. Instead, we use an auxiliary function @lift_nd@ that takes a set and
lifts the elements in the set into the monad stack. Then we can
straightforwardly implement the method @ap_clo@ still in monadic style, where the
closures come from the monads and thus the nondeterminism can be naturally
handled. The light gray line shows what is added from its concrete counterpart.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

\paragraph{Caching and Fixpoint Iteration}
As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called a \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use an @in@ cache and an @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and the @out@ cache is constantly updated during the current iteration. In
the next iteration, the last @out@ cache is served as the @in@ cache, and
an empty cache is plugged-in to the @out@ slot. Once the @out@ cache does not
contain any new information compared with the @in@ cache, the fixed-point is
reached.
In our monad stack, the @in@ and @out@ caches are modeled by the reader monad and
state monad, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}
